\textbf{COMP0085 Summative Assignment}

### 2022-12-30 submission.tex
update gp question

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-grams}
\caption{Covariances for different parameters}
\label{fig:fig2-c-parameter-grams}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
\caption{Samples for different parameters}
\label{fig:fig2-c-parameter-samples}
\end{minipage}
\end{figure}

\begin{itemize}
\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
        This is expected because it is the parameter that adjusts the weight of $\exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right)$.
\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
\item [$\phi$:] As $\phi$ increases, we see less smooth behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
\item [$\eta$:]
\item [$\tau$:]
\item [$\zeta$:]
\end{itemize}
\newpage
\subsection*{(e)}

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
\caption{Without hyperparameter tuning}
\label{fig:f-extrapolation-untrained}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
\caption{With hyperparameter tuning}
\label{fig:f-extrapolation-trained}
\end{minipage}
\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
%\caption{Extrapolation of $CO_2$ concentration levels}
%\label{fig:f-extrapolation-untrained}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
%\caption{Extrapolation of $CO_2$ concentration levels}
%\label{fig:f-extrapolation-trained}
%\end{figure}

@@

>>>>

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
\caption{Samples for different parameters}
\label{fig:fig2-c-parameter-samples}
  \centering
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale = 0.35]{outputs/q2/c-parameter-grams}
\caption{Covariances for different parameters}
\label{fig:fig2-c-parameter-grams}
\end{minipage}
\end{figure}

\begin{itemize}
\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
        This is expected because it is the parameter that adjusts the weight of the periodic component.
\item [$\sigma$:] As $\sigma$ increases, we see reduced periodic behaviour in the sample function.
    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
    This is expected because it adjusts the lengthscale of the periodic portion of the kernel, which ends up dominating the function.
\item [$\phi$:] As $\phi$ increases, we see the ratio of the amplitude of the periodicity component of the sample function reduces compared to the baseline.
    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component.
    This is expected because it adjusts the weight of the non-periodic portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
\item [$\eta$:] As $\eta$ increases we see smoother sample functions.
This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weight of the two components.
\item [$\tau$:] As $\tau$ increases, the period of the periodic function increases.
We can see this reflected in the stripes in the gram matrix getting further apart.
This makes sense because we are adjusting the period in the sinusoid function of the periodic term with $\tau$.
\item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ parameter.
    This places stronger emphasis on the independence of each timestep, which can be seen with the reduction of relative magnitude of off-diagonals in the gram matrix.
    However, this is simply masking the periodic and squared-exponential terms as we can see with the increased magnitude of the functions as $\zeta$ increases.
\end{itemize}
\newpage
\subsection*{(e)}
Suitable values for hyper-parameters can be chosen through a combination of visual inspection and prior knowledge.
For example, it is a reasonable assumption that the $CO_2$ concentration levels have a strong yearly seasonality behaviour due to the cyclic changes in temperature, humidity, etc.
Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this knowledge.
It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
One approach is to maximise:

\[\log P(\textbf{Y} | \textbf{X}) = -\frac{1}{2} \textbf{Y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{Y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]

the log-likelihood  of the posterior distribution with respect to the given data where $\textbf{K}$ is the gram matrix for the kernel (equation 2.30 from http://gaussianprocess.org/gpml/chapters/RW2.pdf).
We can define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal parameters.

Comparing the hyperparameters corresponding to before and after training side by side:

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-untrained-parameters}
\caption{Untrained hyperparameters}
\label{fig:f-untrained-parameters}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-trained-parameters}
\caption{Trained Hyperparmaeters}
\label{fig:f-trained-parameters}
\end{minipage}
\end{figure}

We can analyse some of the changes in these parameters after training to gain some insights.
We can see that $\tau$ remains the same as we would expect given the yearly seasonality we have prior knowledge of.
On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.

\newpage
\subsection*{(f)}
Extrapolating the $CO_2$ concentration levels:

\begin{figure}[h]
\centering
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
\caption{Untrained extrapolation}
\label{fig:f-extrapolation-untrained}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
\caption{Trained extrapolation}
\label{fig:f-extrapolation-trained}
\end{minipage}
\end{figure}

We can see that the extrapolation shows a continued increase in $CO_2$ in the future.
This follows our expectations given that the levels has been steadily increasing in the past.
Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect.
We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the values from before and after training.
Prior to training, the extrapolated prediction is not representative of the given data, with pretty much no seasonal behaviour and very large uncertainty.
After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.

@@

This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a point estimate when making prediction.
For a fully Bayesian approach, we should also incorporate the uncertainty of the linear regression parameters into our extrapolation/uncertainty bounds.
For our procedure, we only include the uncertainty of $g(t)$ however it can be observed in the plots that the trend is not perfectly linear so this should be reflected in the uncertainty of our extrapolation.
Another approach could be to add a linear kernel to our combined kernel function and model $f(t)$ directly with our kernel, removing the linear regression component in our procedure.
Thus our kernel extrapolation would incorporate the uncertainty of all components of our signal.
